Defining Transcriptome using RNA Sequencing\\ of {\em Streptococcus mutans}

### 2012-01-22 package/manuscript.Rnw
Just an upate.

@@

\section{Results}
\comment{
Findings include:
\begin{enumerate}
\item New genes, open reading frames that are not annotated in NCBI's genome database.
\end{enumerate}
}
We sequenced a whole prokaryotic transcriptome of \textit{S.\ mutans}, and
mapped short reads to the genomic sequence of the reference genome UA159
\citep{Ajdic2002}. 
\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
genomic sequence in order to parase \textit{B.\ anthracis} genome into a
sequence of transcripts. Most of annotated genes belong to the predicted
transcripts. We found discrepancies of annotated genes and their transcriptions:
1. A gene contains two transcripts, 
2. Non-coding regions contain transcripts.
\fixme{We could find more discrepancies. Try to find them}. 
Expression levels of adjacent genes within a transcript should be more
correlated to each other than those of adjacent genes belonging to separate transcripts
\citep{Martin2010}.

We identified a set of transcripts, albeit not a full, including small RNAs, and novel
transcripts. 
We hope that our study could serve as a foundation for a comprehensive study of
the \textit{S.\ mutans} transcriptome.
Our current knowledge of the transcriptome is limited to annotated or predicted genes.

Whole transcriptome sequencing using high throughput sequencing (HTS)
technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
transcriptome topography of \textit{S.\ mutans}.

\subsection{From Reviews}
We reconstructed the full-length transcripts by transcriptome assembly using
xxx.

We tried in vain to use short-read assemblers such as Velvet \citep{Zerbino2008} that had been developed to
assembly transcriptome.
Trinity

gene-dense transcriptomes

We ensured a high-quality transcriptome assembly.

@@

We list differentially expressed genes in the comparison of UA159 wild-type and
TW1 mutant-type in Table \ref{tabGlucoseOnly}.  
Of \Sexpr{nrow(counts(smutansGenes))} genes, 
\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
at FDR of 0.1\%.

Table \ref{tabGalactoseOnly}.  
Table \ref{tabUA159Only}.  
Table \ref{tabTW1Only}.

\fixme{Let's compare the microarray result and that from RNA-seq.}
\fixme{Use KEGG for functional category study.}
\fixme{Display results in UCSC genome browser.}
\fixme{Let's find strand-specific RNA-seq.}
\fixme{Add gene descriptions and functional category in the result of DE study.}

We verified the genes using RNA-seq data.

CcpA affects CCR of \textit{fruA}. 
\textit{fruA} fructan hydrolase
When Wen2002 deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
almost completely alleviated. 
What are CCR-sensitive genes? 
What are the mechanism of CcpA on CCR of the \textit{fruA} operon? 
Expression levels of \textit{fruA} in \textit{S.\ mutans} strains that were
grown to early exponential phase (OD600, 0.2) in TV medium supplemented with
0.5\% glucose or galactose. \fixme{What does this mean?} 
Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
\textit{fruA} expression exerted by both glucose and fructose. 
CcpA could directly interact with the \textit{fruA} promoter region.

The role of CcpA in gene expression during growth on glucose. 
\textit{S.\ mutans} UA159 and TW1 grown under catabolite-repressing conditions
(glucose), and catabolite-derepressing conditions (galactose). Under
glucose-only condition I found xx number of differentially expressed genes. What
are the functional categories of those differentially expressed genes? What are
the genes that are most profoundly up- or down-regulated (Smu.1421 to Smu.1424,
PDH enzyme complex), ... and more? ... This regulatory control was
short-circuited in the strain growing on glucose and lacking CcpA. I showed that
some of KEGG pathways were affected by TW1 grown under glucose. We found
down-regulated genes, which would have something to do with pathogenicity. 
Upto here, I compared UA159 and TW1 under glucose.

1. UA159 (A) vs. TW1 (B) : fold change is B/A.
When glucose was the sole carbohydrate for growth, 170 genes were differentially
expressed at least twofold as a result of loss of CcpA (Fig. 2; see Table S1 in
the supplemental material). Of the gene products, 22 were predicted to partici
pate in energy metabolism, 9 were PTS components, and 43 were hypothetical
proteins (Fig. 2). Among the most profoundly up-regulated genes in the ccpA
mutant strain were the genes that encoded the components of the PDH enzyme com-
plex (Smu.1421 to Smu.1424), pyruvate-formate lyase, enzymes in the
tricarboxylic acid (TCA) cycle, selected sugar-specific PTS permeases, and
components of the glycogen biosynthetic machinery (see Table S1 in the
supplemental material). 
Consistent with the profile data, TW1 cells grown in glucose had 78\% higher
PDH activity than the wild-type strain (data not shown). Given the impact on
expression of PDH, TCA enzymes, and selected glycolytic enzymes, CcpA clearly
legislates gene expression to discriminate between carbohydrate-replete and
carbohydrate-limiting conditions. In particular, when there is excess
carbohydrate, S. mutans moves carbohydrates primar- ily through lactate
dehydrogenase, sacrificing an ATP and generating NAD When carbohydrate is
limiting, S. mutans pro- duces mainly formic and acetic acids, shunts
carbohydrate through pyruvate formate lyase, and uses its partial TCA cycle to
regenerate reducing equivalents to maintain an appropriate NAD/NADH balance.
This regulatory control was clearly short-circuited in the strain growing on
glucose and lacking CcpA.
Among the more interesting genes that were down-regu- lated, significant
decreases in expression were noted for the genes encoding components of the
proton-translocating F-ATPase, phosphoglycerate kinase (Pgk), and
fructosyltrans- ferase (ftf), which catalyzes the conversion of sucrose to high-
molecular-weight polymers of fructose, as well as gtfD, which encodes the
glucosyltransferase D enzyme that produces predominantly -1,6-linked polymers
of glucose from sucrose. Since acid tolerance (ATPase), acid production from
carbohy- drate (Pgk), and glucan and fructan production (Gtf and Ftf), as well
as the ability to transport carbohydrates and produce intracellular storage
compounds (glycogen), as noted above (PTS and Glg), are central to the
pathogenic potential of S. mutans, CcpA represents a major control point for
expression of critical virulence attributes in cells presented with glucose.

Growth in derepressing conditions induces major changes in the UA159
transcriptome. Here, I compared glucose and galactose conditions for growing
UA159. Energy metabolism was enriched with differentially expressed genes. There
were more up-regulated genes in galactose condition compared with glucose
condition. Some genes are described in detail.

2. I compared glucose (A) and galactose (B) for growing UA159
It is noteworthy that only three genes in the energy metabolism category were
down-regulated, whereas 23 genes in this category were markedly up-regulated. In
addition, the expression of 28 genes encoding hypothetical proteins was affected,
and 19 of them were up-regulated. The genes encod- ing proteins belonging to the
pathways for lactose and galac- tose utilization (2) were the most strongly
induced genes in cells growing on galactose. The galKTE genes, encoding the
enzymes for galac- tose utilization via the Leloir pathway (2), were
up-regulated between 10- and 36-fold, but the genes of the lactose (lac) operon,
encoding the lactose PTS permease and the tagatose- 6-phosphate pathway, were
up-regulated 70- to 300-fold. Since the products of the galKTE operon contribute
to the genera- tion of UDP-galactose, which is required for cell wall biogen-
esis, the lower level of induction of galKTE than of the lac operon may reflect
a need for higher constitutive levels of expression of the Leloir pathway in the
absence of galactose. Components of the multiple-sugar metabolism operon (msm)
were also induced in cells growing on galactose. The Msm pathway is induced by
raffinose and melibiose, both of which are ?-galactosides (5). The enhanced
expression of gene products involved in en- ergy metabolism, including pyruvate
formate lyase and com- ponents of the TCA cycle, in cells growing on galactose
is consistent with the idea that the cells perceive carbohydrate limitation.

Now, I compared glucose (A) and galactose (B) for growing TW1.

Now, I compared UA159 and TW1 under galactose.

What genes are differentially expressed?

>>>>

Expression levels of \textit{fruA} in \textit{S.\ mutans} strains that were
grown to early exponential phase (OD600, 0.2) in TV medium supplemented with
0.5\% glucose or galactose.

@@

\comment{
Findings include:
\begin{enumerate}
\item New genes, open reading frames that are not annotated in NCBI's genome database.
\end{enumerate}
}

\section{Results}
We sequenced a whole prokaryotic transcriptome of \textit{S.\ mutans}, and
mapped short reads to the genomic sequence of the reference genome UA159
\citep{Ajdic2002}. 
\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
genomic sequence in order to parase \textit{B.\ anthracis} genome into a
sequence of transcripts. Most of annotated genes belong to the predicted
transcripts. We found discrepancies of annotated genes and their transcriptions:
1. A gene contains two transcripts, 
2. Non-coding regions contain transcripts.
\fixme{We could find more discrepancies. Try to find them}. 
Expression levels of adjacent genes within a transcript should be more
correlated to each other than those of adjacent genes belonging to separate transcripts
\citep{Martin2010}. 
We identified a set of transcripts, albeit not a full, including small RNAs, and novel
transcripts. 
We hope that our study could serve as a foundation for a comprehensive study of
the \textit{S.\ mutans} transcriptome.
Our current knowledge of the transcriptome is limited to annotated or predicted genes.
Whole transcriptome sequencing using high throughput sequencing (HTS)
technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
transcriptome topography of \textit{S.\ mutans}. We did not aim at precisely
pin-pointing transcript start or end sites.

We admit that strand-specific RNA-seq would be important in gene-dense
transcriptomes of species like bacteria or archaea.

\subsection{From Reviews}

We could not use short-read assemblers such as Velvet \citep{Zerbino2008} 
and Trinity that had been developed to assembly transcriptome. Because genes in
bacteria are so dense, RNA-seq data without strands did not allow us to apply
any state-of-the-art software of transcriptome assembly.
We ensured a high-quality transcriptome assembly.

@@

We list differentially expressed genes in the comparison of UA159 wild-type and
TW1 mutant-type in Table \ref{tabGlucoseOnly}.  
Of \Sexpr{nrow(counts(smutansGenes))} genes, 
\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
at FDR of 0.1\%. 
We confirmed that the most upregulated genes in the TW1 mutant-type encoded the
components of the PDH enzyme complex (SMU.1421, SMU.1422, SMU.1423, and
SMU.1424).  
Figure \ref{figPieGlucoseOnly} shows the proportions of functional categories 
of the differentially expressed genes. Notably, we found
\Sexpr{length(grep("Energy metabolism",tableGlucoseOnly[,"Functional Class"]))}
genes that were involved in energy metabolism. Sugars and fermentation
subclasses were profound in energy metabolism. We also found 
\Sexpr{length(grep("PTS",tableGlucoseOnly[,"Functional Class"]))}
genes in PTS class of signal transduction. Another subclass, 
``carbohydrates, organic alcohols, and acids,'' in
``transport and binding proteins'' was following next in the number of
differentially expressed genes. We also found  
\Sexpr{length(grep("hypothetical",tableGlucoseOnly[,"Gene Description"]))}
genes annotated as hypothetical, which might constitute the unknown functional
class. These results were consistent with those of \citet{Abranches2008}.
Profoundly upregulated genes in response to the lack of CcpA were
putative Clp proteinase (SMU.1425), 
branched-chain alpha-keto acid dehydrogenase subunit E2 (SMU.1421), and
putative transcriptional regulator (SMU.1599).
Down-regulated genes include cytoplasmic alpha-amylase (SMU.1590), 
beta-D-fructosyltransferase (SMU.2028), and
putative glycerol uptake facilitator protein (SMU.396).
Fructosyltransferase converts sucrose to fructose.

<<echo=FALSE,result=hide>>=
#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
#                              condA="ua159", condB="tw1" )
#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
load("galactoseOnly.RData")
x <- smutans.de2List( galactoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableGalactoseOnly <- x
@
We found that differentially expressed genes in galactose-grown cells were
smaller than those in glucose-grown cells, which was different from the result
of \cite{Abranches2008}. 
Energy metabolism was the most
profoundly differentially expressed (fig.\ \ref{figPieGalactoseOnly}).
Subclasses pyruvate family in amino acid biosynthesis, 
and degradation of proteins, peptides, and glycopeptides in protein fate
were following the next. Almost quarter of the genes were unknown.
Down-regulated genes included cytoplasmic alpha-amylase (SMU.1590), and
beta-D-fructosyltransferase (SMU.2028)
(table \ref{tabGalactoseOnly}).

<<echo=FALSE,result=hide>>=
#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
#ua159only <- smutans.de2( ua159only, type="ua159", 
#                          condA="glucose", condB="galactose" )
#save(ua159only, file=file.path("ua159only.RData"))
load("ua159only.RData")
x <- smutans.de2List( ua159only, qval=1e-10 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableUA159Only <- x
@
Figure \ref{figPieUA159Only} shows proportons of functional categories that
included differentially expressed genes between glucose-grown and
galactose-grown UA159 wild-type cells. Qualitatively, energy metabolism and PTS
classes were profound. This was silmilar to the comparison between UA159
wild-type and TW1 mutant-type grown in glucose. More than quarter of the
differentially expressed genes were unknown, unassigned, or hypothetical.
Consecutive genes (SMU.1488 to SMU.1496) were upregulated in 
catabolite-derepressing conditions or in galactose-grown cells
(\fixme{Are these the genes of the lactose (lac) operon?})
(table \ref{tabUA159Only}).  Another set of genes (SMU.886 to SMU.889) were also
upregulated (\fixme{Are these called galKTE?}).  
Genes from SMU.1535 to SMU.1538 were upregulated. 
Of the \Sexpr{nrow(tableUA159Only)} differentially expressed genes at
$1 \times 10^{-10}$ cutoff of multiple testing adjusted p-value, 
\Sexpr{length(grep("hypothetical",tableUA159Only[,"Gene Description"]))} were
hypothetical, and most of them were down-regulated. 
A few of the down-regulated genes were annotated.

We also found sugar metabolism and PTS signal transduction to be profound in
differential expression in the comparison of glucose-grown and galactose-grown
TW1 mutant-type cells figure \ref{figPieTW1Only}. Many of the genes were also
hypothetical. We found similar genes differentially expressed in TW1 mutant-type
cells to those in UA159 wild-type (table \ref{tabTW1Only}).

\fixme{We may need dot plots where x-axis represents fold changes, and y-axis
functional classes. Dots may be colored by p-values. This would show most of
information in one plot. We could label some of dots to emphasixe. Done?}
\fixme{Use KEGG for functional category study. Done?}
\fixme{Display results in UCSC genome browser. Done?}
\fixme{Use two factor analysis. Done?}

%

Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
\textit{fruA} expression exerted by both glucose and fructose. 
When \citet{Wen2002} deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
almost completely alleviated.  They also showed that 
CcpA directly interacted with the \textit{fruA} promoter region.

The role of CcpA in gene expression during growth on glucose.

\textit{S.\ mutans} UA159 and TW1 grown under catabolite-repressing conditions
(glucose), and catabolite-derepressing conditions (galactose). Under
glucose-only condition we found xx number of differentially expressed genes.
